---
title: "The Tidy R in OMOP CDM ecosystem"
subtitle: "An ecosystem of R packages to work with the OMOP CDM"
format:
  revealjs: 
    theme: [simple, styleWG.scss]
    incremental: true   
    slide-number: true
    chalkboard: true
    preview-links: auto
    margin: 0.07
    code-link: true
    code-line-numbers: false
    height: 900
    width: 1600
    footer: |
      <div style="position: relative; width: 100%;">
        <div style="text-align: center; font-weight: 500;">Tidy R in OMOP CDM ecosystem</div>
        <div style="position: fixed; top: 20px; right: 20px;"><a href="index.qmd" style="text-decoration: none;">⬅ Presentations</a></div>
      </div>
execute:
  echo: true
  eval: true
editor: visual
---

```{r, echo=FALSE}
options(width = 120)
```

```{r, echo=FALSE}
library(dplyr)
library(purrr)
library(jsonlite)
library(patchwork)
library(ggplot2)
library(rlang)
# prepare data
packages <- dplyr::tribble(
  ~package_name, ~classification, ~organisation, ~key_functions,
  "omopgenerics", "Core", "darwin-eu", "Define methods and classes for the ecosystem",
  "CDMConnector", "Back-end", "darwin-eu", "Create a cdm_reference for database backends",
  "omock", "Core", "ohdsi", "Generate mock synthetic data; Create cdm reference to existing mock datasets",
  "visOmopResults", "Core; Visualisation", "darwin-eu", "Manage summarised_result objects; Create tables from summarised_result objects; Create plots from summarised_result objects",
  "CodelistGenerator", "Core; Diagnostics", "darwin-eu", "Manage codelists and concept sets; Run codelist diagnostics",
  "PatientProfiles", "Core", "darwin-eu", "Add columns related to patient demographics; Add columns related to intersections",
  "CohortConstructor", "Analytics", "ohdsi", "Generate base cohorts; Apply inclusion (and exclusion) criteria to cohorts",
  "IncidencePrevalence", "Analytics", "darwin-eu", "Identify denominators based on demographics; Estimate Incidence; Estimate Point-Prevalence; Estimate Period-Prevalence",
  "DrugUtilisation", "Analytics", "darwin-eu", "Estimate drug use; Estimate Proportion of Patients Covered; Summarise Indication; Analyse Discontunation and restart; Calculate daily dose",
  "CohortCharacteristics", "Analytics", "darwin-eu", "Extract counts and attrition of cohorts; Overlap and Timing between cohorts; Characterise cohorts based in demographics and predefined conditions; Data driven characterisation",
  "CohortSurvival", "Analytics", "darwin-eu", "Estimate survival probabilities; Estimate competing risks probabilities",
  "CohortSymmetry", "Analytics", "ohdsi", "Summarise temporal symmetry; Summarise sequence ratios",
  "OmopSketch", "Diagnostics; Analytics", "ohdsi", "Summarise data standards of omop tables; Summarise record trends; Summarise missing data; Summarise concept counts",
  "MeasurementDiagnostics", "Diagnostics", "ohdsi", "Diagnotics for measurement codelists",
  "PhenotypeR", "Diagnostics", "ohdsi", "Diagnostics at database level; Diagnostics at codelist level; Diagnostics at cohort level",
  "OmopViewer", "Visualisation", "ohdsi", "Create automatic shiny app from standard results",
  "DrugExposureDiagnostics", "Diagnostics", "darwin-eu", "Diagnostics for drug codelists",
  #"TreatmentPatterns", "Analytics", "darwin-eu", "Analyse treatment sequences and pathways",
  "OmopConstructor", "ETL", "ohdsi", "Generate the observation_period using cdm data; Run Achilles analyses; Create drug_era table;*Create pregnancy algorithm*",
  "OmopOnSpark", "Back-end", "ohdsi", "Create a cdm_reference using spark connection",
  "OmopOnPostgres", "Back-end", "ohdsi", "Create a cdm_reference using postgres connection; *Use postgres indexes*",
  "OmopOnDuckDB", "Back-end", "ohdsi", "Create a cdm_reference using duckdb connection; *Create a cdm_reference using parquet files*",
  "OmopOnSqlServer", "Back-end", "ohdsi", "Create a cdm_reference using sql server connection;*Create a cdm_reference using synapse connection*"
) |>
  dplyr::left_join(
    tools::CRAN_package_db() |>
      dplyr::as_tibble() |>
      dplyr::select(
        package_name = Package, imports = Imports, suggests = Suggests, 
        last_version = Version, description = Description, title = Title
      ),
    by = "package_name"
  ) |>
  dplyr::mutate(
    key_functions = purrr::map(key_functions, \(kf) {
      paste0("- ", stringr::str_trim(stringr::str_split_1(kf, ";")))
    }),
    github = paste0("www.github.com/", organisation, "/", package_name),
    website = paste0("https://", organisation, ".github.io/", package_name, "/")
  )
pkgs <- packages$package_name

for (pkg in pkgs) {
  if (is.na(packages$imports[packages$package_name == pkg])) {
    desc <- as.list(read.dcf(url(paste0("https://raw.githubusercontent.com/ohdsi/", pkg, "/refs/heads/main/DESCRIPTION")))[1,])
    packages$imports[packages$package_name == pkg] <- desc$Imports %||% "omopgenerics"
    packages$suggests[packages$package_name == pkg] <- desc$Suggests
    packages$description[packages$package_name == pkg] <- desc$Description
    packages$title[packages$package_name == pkg] <- desc$Title
  }
}

packages <- packages |>
  dplyr::mutate(
    last_version = dplyr::coalesce(last_version, "-"),
    imports = purrr::map(imports, \(x) {
      x <- stringr::str_split_1(string = x, pattern = ",") |>
        stringr::str_replace_all("\n", "") |>
        stringr::str_remove("\\(.*") |>
        stringr::str_trim()
      x[x %in% pkgs]
    }),
    suggests = purrr::map(suggests, \(x) {
      x <- stringr::str_split_1(string = x, pattern = ",") |>
        stringr::str_remove("\\(.*") |>
        stringr::str_trim()
      x[x %in% pkgs]
    }),
    imports = dplyr::if_else(
      package_name != "omopgenerics",
      purrr::map(imports, \(x) unique(c("omopgenerics", x))),
      imports
    ),
    logo = map_chr(paste0(organisation, "/", package_name), \(x) {
      url <- paste0("https://github.com/", x, "/blob/main/man/figures/logo.png?raw=true")
      status <- tryCatch(
        httr::status_code(httr::HEAD(url)),
        error = function(e) NA
      )
      if (identical(status, 200L)) {
        url
      } else {
        NA_character_
      }
    })
  )

versions <- pkgs |>
  purrr::map(\(pkg) {
    url <- paste0("https://crandb.r-pkg.org/", pkg, "/all")
    status <- tryCatch(
      httr::status_code(httr::HEAD(url)),
      error = function(e) NA
    )
    
    if (identical(status, 200L)) {
      
      cli::cli_inform("Getting releases for {pkg}")
      
      res <- jsonlite::fromJSON(url) |>
        purrr::pluck("versions") |>
        purrr::map_dfr(~ dplyr::tibble(
          package_name = pkg,
          version = .$Version,
          date = as.Date(.$Date)
        )) |> 
        dplyr::arrange(date) |> 
        dplyr::mutate(
          version = dplyr::if_else(
            package_name == "CodelistGenerator" & 
              version == "3.0",
            "3.0.0",
            version)) |> 
        dplyr::mutate(
          first = as.integer(gsub("^(\\d+)\\..*$", "\\1", version)),
          second = as.integer(gsub("^\\d+\\.(\\d+)\\..*$", "\\1", version)),
          third = as.integer(gsub("^.*\\.([0-9]+)$", "\\1", version))
        ) |> 
        dplyr::mutate(previous_first = dplyr::lag(first),
                      previous_second = dplyr::lag(second),
                      previous_third = dplyr::lag(third)) |> 
        dplyr::mutate(release_type = dplyr::case_when(
          is.na(previous_first) ~ "first release",
          first > previous_first  ~ "major release",
          second > previous_second  ~ "minor release",
          third > previous_third  ~ "patch release",
          .default = NA
        )) |> 
        dplyr::select(package_name, version, date, release_type)
      
    } else {
      res <- NULL
    }
    
    res
  }) |>
  purrr::compact() |>
  dplyr::bind_rows() |>
  mutate(package_name = factor(.data$package_name, levels = rev(packages$package_name)))

packages <- packages |>
  dplyr::left_join(
    versions |>
      group_by(package_name) |>
      summarise(first_release = min(date), n_releases = dplyr::n()),
    by = "package_name"
  ) |>
  dplyr::mutate(
    first_release = dplyr::coalesce(as.character(first_release), "-"),
    n_releases = dplyr::coalesce(n_releases, 0)
  )

ranges <- packages |>
  dplyr::select(package_name, organisation) |>
  purrr::pmap(\(package_name, organisation) {
    x <- gh::gh("/repos/{owner}/{repo}", owner = organisation, repo  = package_name)
    dplyr::tibble(
      package_name = package_name,
      min_date = min(as.Date(x$created_at), min(versions$date[versions$package_name == package_name])),
      max_date = Sys.Date()
    )
  }) |>
  dplyr::bind_rows() |>
  mutate(package_name = factor(.data$package_name, levels = rev(packages$package_name)))

colors <- c(
  "Core" = "#e3f2fd",
  "Back-end" = "#e8f5e9",
  "Analytics" = "#fff3e0",
  "Diagnostics" = "#fce4ec",
  "ETL" = "#ede7f6",
  "Visualisation" = "#e0f7fa"
)
```

## Content

- Motivation and foundations
- Existing packages overview
- `omopgenerics` the core package

## Motivation

-   OMOP CDM
-   Complications to install packages, not in cran, java, contribute to the ecosystem
-   Tidyverse
-   No Java code, no SQL code, just R code

## Foundation

```{r}
library(palmerpenguins)
library(dplyr)

penguins |>
  group_by(species) |>
  summarise(
    n = n(),
    min_bill_length_mm = min(bill_length_mm, na.rm = TRUE),
    mean_bill_length_mm = mean(bill_length_mm, na.rm = TRUE),
    max_bill_length_mm = max(bill_length_mm, na.rm = TRUE)
  ) |>
  mutate(min_max_bill_length_mm = paste0(
    min_bill_length_mm, " to ", max_bill_length_mm
  )) |>
  select("species", "mean_bill_length_mm", "min_max_bill_length_mm")
```

## Foundartion

```{r}
library(duckdb)

con <- dbConnect(drv = duckdb())
dbWriteTable(conn = con, name = "penguins", value = penguins)
penguins_db <- tbl(con, "penguins")
penguins_db
```

## Foundartion

```{r}
penguins_db |>
  group_by(species) |>
  summarise(
    n = n(),
    min_bill_length_mm = min(bill_length_mm, na.rm = TRUE),
    mean_bill_length_mm = mean(bill_length_mm, na.rm = TRUE),
    max_bill_length_mm = max(bill_length_mm, na.rm = TRUE)
  ) |>
  mutate(min_max_bill_length_mm = paste0(
    min_bill_length_mm, " to ", max_bill_length_mm
  )) |>
  select("species", "mean_bill_length_mm", "min_max_bill_length_mm")
```


## Foundation

```{r}
penguins_db |>
  group_by(species) |>
  summarise(
    n = n(),
    min_bill_length_mm = min(bill_length_mm, na.rm = TRUE),
    mean_bill_length_mm = mean(bill_length_mm, na.rm = TRUE),
    max_bill_length_mm = max(bill_length_mm, na.rm = TRUE)
  ) |>
  mutate(min_max_bill_length_mm = paste0(
    min_bill_length_mm, " to ", max_bill_length_mm
  )) |>
  select("species", "mean_bill_length_mm", "min_max_bill_length_mm") |>
  show_query()
```

## Foundation

[dbplyr](https://dbplyr.tidyverse.org/) support the database translation of [dplyr](https://dplyr.tidyverse.org/) functions against several database management systems.

. . .

The presented ecosystem sits on top of these functions combined with the known structure of the [OMOP CDM](https://ohdsi.github.io/CommonDataModel/) tables to build packages to perform Epidemiological studies extracting data from an OMOP CDM source that can be in several different forms.

## Cronology (Early stages)

```{r, echo=FALSE, message=FALSE, warning=FALSE}
plotCronology <- function(endDate) {
  endDate <- as.Date(endDate)
  p1 <- versions |>
    filter(date <= endDate) |>
    ggplot() +
    geom_segment(
      data = ranges |>
        mutate(
          min_date = if_else(min_date < endDate, min_date, endDate),
          max_date = if_else(max_date < endDate, max_date, endDate)
        ),
      aes(x = min_date, xend = max_date, y = package_name, yend = package_name),
      color = "grey90", 
      linewidth = 1.5
    ) +
    geom_point(aes(date, package_name,
                   fill = release_type,
                   size = release_type),
               shape = 21) +
    scale_size_manual(
      values = c("first release" = 3, 
                 "patch release" = 1, 
                 "minor release" = 2, 
                 "major release" = 3)
    ) +
    scale_fill_manual(
      values = c("first release" = "#e41a1c", 
                 "patch release" = "#377eb8", 
                 "minor release" = "#4daf4a", 
                 "major release" = "#984ea3")
    ) +
    theme_bw() +
    ylab("") +
    xlab("") +
    # facet_grid(type~., scales = "free_y") +
    theme(legend.title = element_blank(),
          legend.position = "top",
          panel.spacing = unit(0, "lines"))
  p1 <- p1+ theme(
    axis.text.x = element_blank(), 
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank()) 
  
  p2 <- versions |> 
    filter(date <= endDate) |>
    mutate(date_floor = lubridate::floor_date(date, unit = "month")) |> 
    group_by(date_floor) |> 
    tally() |> 
    ggplot() +
    geom_col(aes(date_floor, n), width = 30, colour = "black", fill = "grey70")+
    theme_bw() +
    ylab("") +
    xlab("Releases in month")
  
  rng <- c(min(ranges$min_date), max(ranges$max_date))
  
  p1 <- p1 +
    scale_x_date(date_breaks = "1 year", date_labels = "%Y", limits = rng)
  
  p2 <- p2 +
    scale_x_date(date_breaks = "1 year", date_labels = "%Y", limits = rng)
  
  p1 / p2 +
    plot_layout(heights = c(4, 1), guides = "collect") &
    theme(axis.title.x = element_blank(), legend.position = "top")
}
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
plotCronology("2023-01-01")
```

## Cronology (First packages)

```{r, echo=FALSE, message=FALSE, warning=FALSE}
plotCronology("2024-01-01")
```

## Cronology (Standardarisation and expansion)

```{r, echo=FALSE, message=FALSE, warning=FALSE}
plotCronology("2025-01-01")
```

## Cronology (1.0.0 releases and stability)

```{r, echo=FALSE, message=FALSE, warning=FALSE}
plotCronology("2026-01-01")
```

## Existing packages

:::: {.card-grid}
```{r, echo=FALSE, results='asis'}
xx <- packages |>
  mutate(
    classification = map(classification, \(x) {
      stringr::str_split_1(x, ";") |>
        stringr::str_trim()
    })
  ) |>
  tidyr::unnest("classification") |>
  group_by(package_name) |>
  mutate(italic = if_else(n() > 1, 1, 0)) |>
  ungroup() |>
  group_by(classification) |>
  group_split() |>
  as.list()
names(xx) <- map_chr(xx, \(x) unique(x$classification))
xx <- xx[names(colors)]
xx |>
  map(\(x) {
    nm <- unique(x$classification)
    c(
      paste0('::: {.pkg-card style="background:', colors[nm], '"}'),
      paste0("**", nm, "**"),
      paste0("- ", if_else(x$italic == 1, "*", ""), x$package_name, if_else(x$italic == 1, "*", "")),
      ":::"
    ) |>
      paste0(collapse = "\n\n")
  }) |>
  paste0(collapse = "\n\n") |>
  cat()
```
::::

```{r, results='asis', echo=FALSE, warning=FALSE, message=FALSE}
packages |>
  dplyr::select(package_name, classification, last_version, title, description, website, github, n_releases, first_release, key_functions, logo) |>
  purrr::pmap(\(package_name, classification, last_version, title, description, website, github, n_releases, first_release, key_functions, logo) {
    c(
      paste0("## ", package_name, " (", classification, ")"),
      if_else(is.na(logo), "", paste0(
        '\n<div style="position:absolute; top:50px; right:20px;"><img src="', logo, '" width="80"></div>\n'
      )),
      ". . .",
      "",
      paste0("**", title, "**"),
      "",
      paste0("Current version: **", last_version, "** First release: *", 
             first_release, "* Total releases: *", n_releases, "*"),
      "",
      paste0("Links: [Website](", website, ");  [GitHub](", github, ")"),
      "",
      ". . .",
      "",
      paste0("*", description, "*"),
      "",
      key_functions,
      ""
    ) |>
      paste0(collapse = "\n")
  }) |>
  paste0(collapse = "\n\n") |>
  cat()
```

## How packages interact?

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(igraph)
library(ggraph)
library(tidygraph)
library(tools)
library(jsonlite)
library(dplyr)
library(purrr)
library(patchwork)
library(ggplot2)
library(cranlogs)
library(bsicons)
library(ggdag)

dependencies <- packages |>
  select(package_name, imports) |>
  tidyr::unnest("imports") |>
  rename(from = package_name, to = imports) |>
  mutate(type = "Imports") |>
  union_all(
    packages |>
      select(package_name, suggests) |>
      tidyr::unnest("suggests") |>
      rename(from = package_name, to = suggests) |>
      mutate(type = "Suggests")
  )

plot_dependencies <- function(dependencies) {
  g <- as_tbl_graph(graph_from_data_frame(dependencies, directed = TRUE, vertices = pkgs))
  g <- g %>%
    mutate(
      dependencies = centrality_degree(mode = "in")
    )
  ggraph(g, layout = "centrality", centrality = dependencies) +
    geom_edge_arc(aes(linetype = type, color = type),
                  strength = 0.1, end_cap = circle(4, 'mm'),
                  arrow = arrow(length = unit(3, 'mm'), type = "closed")) +
    geom_node_point(aes(fill = dependencies),
                    shape = 21, colour = "black",
                    size = 5) +
    geom_node_text(aes(label = name),
                   repel = TRUE,
                   bg.color = "white",
                   bg.r = 0.15,
                   size = 4) +
    scale_edge_linetype_manual(name = "Dependency type",
                               values = c("Imports" = "solid",
                                          "Suggests" = "dashed")) +
    scale_edge_color_manual(name = "Dependency type",
                            values = c("Imports" = "black",
                                       "Suggests" = "grey45")) +
    scale_fill_viridis(name = "Reverse dependencies",
                       option="mako",
                       direction = -1) +
    theme_void() +
    theme(legend.position = "bottom",
          legend.box = "veritcal") 
}
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
dependencies |>
  filter(type == "Imports") |>
  plot_dependencies()
```

## How packages interact?

```{r, echo=FALSE, message=FALSE, warning=FALSE}
dependencies |>
  plot_dependencies()
```

## Some statistics

```{r, echo=FALSE, message=FALSE, warning=FALSE}
downloads <- cranlogs::cran_downloads(
  packages = packages$package_name, 
  from = as.Date("2022-01-01"), 
  to = Sys.Date()
) 

total_per_package <- downloads |> 
  group_by(package) |> 
  summarise(total = sum(count)) |> 
  ungroup()

total <- total_per_package |> 
  summarise(total = sum(total))

bslib::value_box(
  title = "Number of R packages",
  value = nrow(packages),
  showcase = bsicons::bs_icon("bar-chart")
)

bslib::value_box(
  title = "Number of R packages on CRAN",
  value = sum(total_per_package$total > 0),
  showcase = bsicons::bs_icon("bar-chart")
)

bslib::value_box(
  title = "Number Releases",
  value = versions |> nrow(),
  showcase = bsicons::bs_icon("bar-chart")
)

bslib::value_box(
  title = "Total CRAN downloads",
  value = total |> pull(),
  showcase = bsicons::bs_icon("bar-chart")
)
```

# omopgenerics

## omopgenerics

Provides definitions of core **classes** and **methods** used by analytic pipelines that query the OMOP (Observational Medical Outcomes Partnership) common data model.

## Classes

- The CDM refernce object: `cdm_reference`, `cdm_source`, `local_source`
- CDM tables: `cdm_table`, `omop_table`, `achilles_table`, `cohort_table`
- Code lists: `codelist`, `codelist_with_details`, `concept_set_expression`
- Result model: `summaried_result`

## `cdm_reference`

The `cdm_reference` is the center of the ecosystem, it is a list with a reference to the different OMOP CDM tables. In general we will create a `cdm_reference` with the *CDMConnector* package:

```{r, eval= FALSE}
library(CDMConnector)
cdm <- cdmFromCon(
  con = con,
  cdmSchema = "public",
  writeSchema = "results",
  writePrefix = "mc_",
  cdmName = "my OMOP"
)
cdm
```
```{r, echo=FALSE}
cdm <- omock::mockCdmFromDataset(source = "duckdb")
attr(cdm, "cdm_name") <- "my OMOP"
```
```{r, echo=FALSE,message=TRUE}
cdm
```

## `cdm_reference`

The `cdm_reference` object contains several attributes:

```{r}
library(omopgenerics)
cdmName(cdm)
```

```{r}
cdmVersion(cdm)
```

```{r, message=TRUE}
cdmSource(cdm)
```

```{r}
sourceType(cdm)
```

## `cdm_reference`

You can use the `cdm_reference` object to insert, list, drop or read tables in using the same source of the `cdm_reference`.

. . .

```{r}
cdm <- insertTable(cdm = cdm, name = "my_table", table = cars)
cdm$my_table
```

. . .

```{r}
listSourceTables(cdm = cdm)
```

. . .

```{r, message=TRUE}
cdm$my_table <- NULL
cdm
```

. . .

```{r, message=TRUE}
cdm <- readSourceTable(cdm = cdm, name = "my_table")
cdm
```

. . .

```{r, message=TRUE}
cdm <- dropSourceTable(cdm = cdm, name = "my_table")
cdm
listSourceTables(cdm = cdm)
```

## `cdm_source`

The `cdm_source` is an object that defines how to:

- compute a table (`compute.<class>`)
- eliminate a table (`dropSourceTable.<class>`)
- insert a table (`insertTable.<class>`)
- read a table (`readSourceTable.<class>`)

. . .

At the creation of the `cdm_source` object all these steps will be checked (that's why creating a `cdm_reference` can take a few seconds).

. . .

To create a new `cdm_source` object you can use `newCdmSource()`.

. . .

For example for CDMConnector ths source contains the connection and the writeSchema:

```{r}
library(omock)
library(omopgenerics)

cdm <- mockCdmFromDataset(source = "duckdb")
src <- cdmSource(cdm)
src
unclass(src)
```

## `cdm_source`

You can summarise the properties of a `cdm_source` object using `summary()`

. . .

```{r}
summary(src)
```

## `db_source`

For example here we can see how the `db_source` is defined by *CDMConnector*: <https://github.com/darwin-eu/CDMConnector/blob/main/R/dbSource.R>

## `cdm_table`

A `cdm_table` object is a table that comes from a `cdm_reference`:

. . .

```{r}
class(cdm$person)
```

. . . 

The `cdm_table` will always include the reference where they come from, so we will be able to access other tables of the same `cdm_reference` object:

```{r, message=TRUE}
cdmReference(cdm$person)
```

## `cdm_table`

The fact that contains the `cdm_refernce` object allow us to also access the same properties that we had before:

```{r}
cdmName(cdm$person)
```

```{r}
cdmVersion(cdm$person)
```

```{r, message=TRUE}
cdmSource(cdm$person)
```

```{r}
sourceType(cdm)
```

## `cdm_table`

On top of that extra attributes can be obtained:

```{r}
tableName(cdm$person)
```

```{r, message=TRUE}
tableSource(cdm$person)
```

## `omop_table`

The `omop_table` is a subclass of `cdm_table` that protects the OMOP CDM integrity and validates the OMOP CDM standard tables.

. . .

```{r}
class(cdm$person)
```

. . .

For example we can not remove required columns from `omop_tables`:

```{r, error=TRUE}
cdm$person <- cdm$person |>
  select(!"person_id")
```

. . .

Or assign tables that are not named the same way:

```{r, error=TRUE}
cdm$person <- cdm$drug_exposure
```

. . .

But this does not mean that we can not modify them, we can add columns and/or change its content:

```{r}
cdm$person <- cdm$person |>
  filter(person_id == 6) |>
  mutate(extra_col = "my_content")
cdm$person |>
  glimpse()
```

## `achilles_table`

The class `achilles_table` is a similar protection for the tables coming form the achilles analyses.

## `cohort_table`

```{r, echo=FALSE}
cdm <- mockCdmFromDataset(source = "duckdb")
cdm$my_cohort <- CohortConstructor::conceptCohort(
  cdm = cdm, 
  name = "my_cohort", 
  conceptSet = list(cohort1 = 40481087L, cohort2 = 4112343L)
)
```

The `cohort_table` class is a class built on top of `cdm_table` (it has all the functionalities seen for `cdm_table`) with some extra attributes.

```{r, message=TRUE}
cdm
```

. . .

```{r}
cdm$my_cohort
```

. . .

```{r}
class(cdm$my_cohort)
```

## `cohort_table`

```{r}
attributes(cdm$my_cohort)
```

## `cohort_table`

We can access those attributes:

. . .

```{r}
settings(cdm$my_cohort)
```

. . .

```{r}
cohortCount(cdm$my_cohort)
```

## `cohort_table`

```{r}
attrition(cdm$my_cohort)
```

## `cohort_table`

We can easily update the attrition as:

```{r}
cdm$my_cohort <- cdm$my_cohort |>
  filter(subject_id %% 2 == 0) |>
  compute(name = "my_cohort") |>
  recordCohortAttrition("Onlye even subjects")
```

. . .

```{r}
cohortCount(cdm$my_cohort)
attrition(cdm$my_cohort)
```

## Codelists

The package defines 3 different ways to define code lists:

- `codelist`
- `codelist_with_details`
- `concept_set_expression`

## `codelist`

```{r, message=TRUE}
codelist <- newCodelist(list(my_codes = c(1, 2, 3), condition2 = c(3, 4)))
codelist
```

```{r}
codelist$my_codes
```

## `codelist_with_details`

```{r, message=TRUE}
codelist <- list(
  my_codes = tibble(
    concept_id = c(1L, 2L, 3L),
    extra_properties = c("a", "b", "c")
  ), 
  condition2 = tibble(
    concept_id = c(3L, 4L),
    concept_name = c("code 3", "code 4")
  )
) |>
  newCodelistWithDetails()
codelist
```

```{r}
codelist$my_codes
```

## `codelist`
## `codelist`

```{r, message=TRUE}
codelist <- newCodelist(list(my_codes = c(1, 2, 3), condition2 = c(3, 4)))
codelist
```

```{r}
codelist$my_codes
```

## `concept_set_expression`

```{r, message=TRUE}
codelist <- list(
  my_codes = tibble(
    concept_id = c(1L, 2L, 3L),
    descendants = c(T, F, T),
    excluded = c(F, T, F),
    mapped = F
  ), 
  condition2 = tibble(
    concept_id = c(3L, 4L),
    descendants = c(T, F),
    excluded = c(F, T),
    mapped = F
  )
) |>
  newConceptSetExpression()
codelist
```

```{r}
codelist$my_codes
```

## Combine and Subset codelists

*omopgenerics* contains methods so that combining and subsetting codelists will preserve the class object:

. . .

```{r}
codelist1 <- newCodelist(list(concept_1 = 1L, anxiety = c(4L, 5L)))
codelist2 <- newCodelist(list(hypertension = 6L))
c(codelist1, codelist2)
```

. . .

```{r}
codelist1["anxiety"]
```

## CodelistGenerator

[CodelistGenerator](https://darwin-eu.github.io/CodelistGenerator/) provides multiple functionality to manage the 3 codelists classes, combine them, compare them:

- `get*Codes()` to create different code lists
- `subset*()` to subset code lists
- `stratify*()` to stratify code lists
- `asCodelist()`, `asCodelistWithDetails()`, `asConceptSetExpression()`

## The conceptSet argument

In general analytical packages will have an argument named `conceptSet` the input of this argument can either be a `codelist`, `codelist_with_details` or `concept_set_expression` or an object that can be converted to any of these objects (e.g. `list(concept = 1)`).

## Export codelist

`codelist` and `concept_set_expressions` can be exported in `json` or `csv` format:

. . .

```{r, eval=FALSE}
library(here)
exportCodelist(codelist, path = here(), type = "csv")
exportCodelist(codelist, path = here(), type = "json")
exportConceptSetExpression(codelist, path = here(), type = "csv")
exportConceptSetExpression(codelist, path = here(), type = "json")
```

. . .

Then later can be also imported:

```{r, eval=FALSE}
codelist <- importCodelist(path = here(), type = "csv")
codelist <- importCodelist(path = here(), type = "json")
codelist <- importConceptSetExpression(path = here(), type = "csv")
codelist <- importConceptSetExpression(path = here(), type = "json")
```

**JSON file definition is the same than ATLAS and in line with the guidelines of OHDSI TAB.**

## `summarised_result`

The `summarised_result` is the result model defined by *omopgenerics*

## Visualisations with `summarised_result`

The [visOmopResults](https://darwin-eu.github.io/visOmopResults/) provides multiple functionality to to create tables and plot from the `summarised_result` object:

- [Create Tables using `summarised_results`](https://darwin-eu.github.io/visOmopResults/articles/a01_tables.html)
- [Create Plots using `summarised_results`](https://darwin-eu.github.io/visOmopResults/articles/a02_plots.html)

## Further reading

- Tidy R packages foundations: <https://ohdsi.github.io/Tidy-R-programming-with-OMOP/>
- Tidy R packages develioment: <https://oxford-pharmacoepi.github.io/Tidy-R-packages-with-OMOP/>
