---
title: "The Tidy R in OMOP CDM ecosystem"
subtitle: "An ecosystem of R packages to work with the OMOP CDM"
format:
  revealjs: 
    theme: [simple, styleWG.scss]
    incremental: true   
    slide-number: true
    chalkboard: true
    preview-links: auto
    margin: 0.07
    code-link: true
    code-line-numbers: false
    height: 900
    width: 1600
    footer: |
      <div style="position: relative; width: 100%;">
        <div style="text-align: center; font-weight: 500;">Tidy R in OMOP CDM ecosystem</div>
        <div style="position: fixed; top: 20px; right: 20px;"><a href="index.qmd" style="text-decoration: none;">⬅ Presentations</a></div>
      </div>
execute:
  echo: true
  eval: true
editor: visual
---

```{r, echo=FALSE}
options(width = 120)
```

```{r, echo=FALSE}
library(dplyr)
library(purrr)
library(jsonlite)
library(patchwork)
library(ggplot2)
library(rlang)
# prepare data
packages <- dplyr::tribble(
  ~package_name, ~classification, ~organisation, ~key_functions,
  "omopgenerics", "Core", "darwin-eu", "Define methods and classes for the ecosystem",
  "CDMConnector", "Back-end", "darwin-eu", "Create a cdm_reference for database backends",
  "omock", "Core", "ohdsi", "Generate mock synthetic data; Create cdm reference to existing mock datasets",
  "OmopConstructor", "ETL", "ohdsi", "Generate the obsercation_period using cdm data; Run Achilles analyses; Create drug_era table;*Create pregnancy algorithm*",
  "OmopOnSpark", "Back-end", "ohdsi", "Create a cdm_reference using spark connection",
  "OmopOnPostgres", "Back-end", "ohdsi", "Create a cdm_reference using postgres connection; *Use postgres indexes*",
  "OmopOnDuckDB", "Back-end", "ohdsi", "Create a cdm_reference using duckdb connection; *Create a cdm_reference using parquet files*",
  "OmopOnSqlServer", "Back-end", "ohdsi", "Create a cdm_reference using sql server connection;*Create a cdm_reference using synapse connection*",
  "visOmopResults", "Core;Visualisation", "darwin-eu", "Manage summarised_result objects; Create tables from summarised_result objects; Create plots from summarised_result objects",
  "CodelistGenerator", "Core;Diagnostics", "darwin-eu", "Manage codelists and concept sets; Run codelist diagnostics",
  "PatientProfiles", "Core", "darwin-eu", "Add columns related to patient demographics; Add columns related to intersections",
  "CohortConstructor", "Analytics", "ohdsi", "Generate base cohorts; Apply inclusion (and exclusion) criteria to cohorts",
  "IncidencePrevalence", "Analytics", "darwin-eu", "Identify denominators based on demographics; Estimate Incidence, Point-Prevalence and Period-Prevalence",
  "DrugUtilisation", "Analytics", "darwin-eu", "Estimate drug use, Proportion of Patients Covered, Indication, Discontunation and restart; Calculate daily dose",
  "CohortCharacteristics", "Analytics", "darwin-eu", "Extract counts and attrition of cohorts; Overlap and Timing between cohorts; Characterise cohorts based in demographics and predefined conditions; Data driven characterisation",
  "CohortSurvival", "Analytics", "darwin-eu", "",
  "CohortSymmetry", "Analytics", "ohdsi", "",
  "OmopSketch", "Diagnostics;Analytics", "ohdsi", "",
  "MeasurementDiagnostics", "Diagnostics", "ohdsi", "",
  "PhenotypeR", "Diagnostics", "ohdsi", "",
  "OmopViewer", "Visualisation", "ohdsi", "",
  "DrugExposureDiagnostics", "Diagnostics", "darwin-eu", "",
  "TreatmentPatterns", "Analytics", "darwin-eu", ""
) |>
  dplyr::left_join(
    tools::CRAN_package_db() |>
      dplyr::as_tibble() |>
      dplyr::select(
        package_name = Package, imports = Imports, suggests = Suggests, 
        last_version = Version, description = Description, title = Title
      ),
    by = "package_name"
  ) |>
  dplyr::mutate(
    key_functions = purrr::map(key_functions, \(kf) {
      paste0("- ", stringr::str_trim(stringr::str_split_1(kf, ";")))
    }),
    github = paste0("www.github.com/", organisation, "/", package_name),
    website = paste0("https://", organisation, ".github.io/", package_name, "/")
  )
pkgs <- packages$package_name

for (pkg in pkgs) {
  if (is.na(packages$imports[packages$package_name == pkg])) {
    desc <- as.list(read.dcf(url(paste0("https://raw.githubusercontent.com/ohdsi/", pkg, "/refs/heads/main/DESCRIPTION")))[1,])
    packages$imports[packages$package_name == pkg] <- desc$Imports %||% "omopgenerics"
    packages$suggests[packages$package_name == pkg] <- desc$Suggests
    packages$description[packages$package_name == pkg] <- desc$Description
    packages$title[packages$package_name == pkg] <- desc$Title
  }
}

packages <- packages |>
  dplyr::mutate(
    last_version = dplyr::coalesce(last_version, "-"),
    imports = purrr::map(imports, \(x) {
      x <- stringr::str_split_1(string = x, pattern = ",") |>
        stringr::str_replace_all("\n", "") |>
        stringr::str_remove("\\(.*") |>
        stringr::str_trim()
      x[x %in% pkgs]
    }),
    suggests = purrr::map(suggests, \(x) {
      x <- stringr::str_split_1(string = x, pattern = ",") |>
        stringr::str_remove("\\(.*") |>
        stringr::str_trim()
      x[x %in% pkgs]
    }),
    imports = dplyr::if_else(
      package_name != "omopgenerics",
      purrr::map(imports, \(x) unique(c("omopgenerics", x))),
      imports
    )
  )

versions <- pkgs |>
  purrr::map(\(pkg) {
    url <- paste0("https://crandb.r-pkg.org/", pkg, "/all")
    status <- tryCatch(
      httr::status_code(httr::HEAD(url)),
      error = function(e) NA
    )
    
    if (identical(status, 200L)) {
      
      cli::cli_inform("Getting releases for {pkg}")
      
      res <- jsonlite::fromJSON(url) |>
        purrr::pluck("versions") |>
        purrr::map_dfr(~ dplyr::tibble(
          package_name = pkg,
          version = .$Version,
          date = as.Date(.$Date)
        )) |> 
        dplyr::arrange(date) |> 
        dplyr::mutate(
          version = dplyr::if_else(
            package_name == "CodelistGenerator" & 
              version == "3.0",
            "3.0.0",
            version)) |> 
        dplyr::mutate(
          first = as.integer(gsub("^(\\d+)\\..*$", "\\1", version)),
          second = as.integer(gsub("^\\d+\\.(\\d+)\\..*$", "\\1", version)),
          third = as.integer(gsub("^.*\\.([0-9]+)$", "\\1", version))
        ) |> 
        dplyr::mutate(previous_first = dplyr::lag(first),
                      previous_second = dplyr::lag(second),
                      previous_third = dplyr::lag(third)) |> 
        dplyr::mutate(release_type = dplyr::case_when(
          is.na(previous_first) ~ "first release",
          first > previous_first  ~ "major release",
          second > previous_second  ~ "minor release",
          third > previous_third  ~ "patch release",
          .default = NA
        )) |> 
        dplyr::select(package_name, version, date, release_type)
      
    } else {
      res <- NULL
    }
    
    res
  }) |>
  purrr::compact() |>
  dplyr::bind_rows() |>
  mutate(package_name = factor(.data$package_name, levels = rev(packages$package_name)))

packages <- packages |>
  dplyr::left_join(
    versions |>
      group_by(package_name) |>
      summarise(first_release = min(date), n_releases = dplyr::n()),
    by = "package_name"
  ) |>
  dplyr::mutate(
    first_release = dplyr::coalesce(as.character(first_release), "-"),
    n_releases = dplyr::coalesce(n_releases, 0)
  )

ranges <- packages |>
  dplyr::select(package_name, organisation) |>
  purrr::pmap(\(package_name, organisation) {
    x <- gh::gh("/repos/{owner}/{repo}", owner = organisation, repo  = package_name)
    dplyr::tibble(
      package_name = package_name,
      min_date = min(as.Date(x$created_at), min(versions$date[versions$package_name == package_name])),
      max_date = Sys.Date()
    )
  }) |>
  dplyr::bind_rows()
```

## Content

- Motivation and foundations
- Existing packages overview
- omopgenerics the core package

## Motivation

-   OMOP CDM
-   complications to install packages, not in cran, java, contribute to the ecosystem
-   tidyverse
-   no Java code, no SQL code, just R code

## Foundation

- dplyr + dbplyr examples

## Create a cdm_reference

## Cronology

```{r, echo=FALSE, message=FALSE, warning=FALSE}
p1 <- versions |>
  ggplot() +
  geom_segment(data = ranges, 
               aes(x = min_date, xend = max_date,
                   y = package_name,
                   yend = package_name),
               color = "grey90", linewidth = 1.5) +
  geom_point(aes(date, package_name,
                 fill = release_type,
                 size = release_type),
             shape = 21) +
  scale_size_manual(
    values = c("first release" = 3, 
               "patch release" = 1, 
               "minor release" = 2, 
               "major release" = 3)
  ) +
  scale_fill_manual(
    values = c("first release" = "#e41a1c", 
               "patch release" = "#377eb8", 
               "minor release" = "#4daf4a", 
               "major release" = "#984ea3")
  ) +
  theme_bw() +
  ylab("") +
  xlab("") +
  # facet_grid(type~., scales = "free_y") +
  theme(legend.title = element_blank(),
        legend.position = "top",
        panel.spacing = unit(0, "lines"))
p1 <- p1+ theme(
  axis.text.x = element_blank(), 
  axis.title.x = element_blank(),
  axis.ticks.x = element_blank()) 

p2 <- versions |> 
  mutate(date_floor = lubridate::floor_date(date, unit = "month")) |> 
  group_by(date_floor) |> 
  tally() |> 
  ggplot() +
  geom_col(aes(date_floor, n), width = 30, colour = "black", fill = "grey70")+
  theme_bw() +
  ylab("") +
  xlab("Releases in month")

rng <- c(min(ranges$min_date), max(ranges$max_date))

p1 <- p1 +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y", limits = rng)

p2 <- p2 +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y", limits = rng)

p1 / p2 +
  plot_layout(heights = c(4, 1), guides = "collect") &
  theme(axis.title.x = element_blank(), legend.position = "top")
```

## Existing packages

```{r, echo=FALSE, results='asis'}
library(dplyr)
library(tidyr)
library(purrr)
library(htmltools)

# your existing packages tibble must be present in the R session
card_colors <- c(
  "Core" = "#e3f2fd",
  "Back-end" = "#e8f5e9",
  "Analytics" = "#fff3e0",
  "Diagnostics" = "#fce4ec",
  "ETL" = "#ede7f6",
  "Visualisation" = "#e0f7fa"
)

# Build a grid of simple cards (no bslib)
cards_html <- tibble(classification = names(card_colors)) |>
  left_join(
    packages %>%
      separate_rows(classification, sep = ";") %>%
      mutate(classification = trimws(classification)) %>%
      group_by(classification) %>%
      summarise(packages = list(sort(package_name)), .groups = "drop"),
    by = "classification"
  ) %>%
  pmap(function(classification, packages) {
    # fallback color in case classification not in palette
    bg <- card_colors[[classification]] %||% "#f5f5f5"
    tags$div(
      class = "fragment pkg-card",
      `data-fragment-index` = NA, # optional: keep DOM order
      style = paste0(
        "background:", bg, ";",
        "border-radius:8px;",
        "padding:0.8rem;",
        "box-shadow: 0 2px 6px rgba(0,0,0,0.06);",
        "min-height: 6rem;"
      ),
      tags$div(class = "pkg-card-title", classification),
      tags$ul(
        style = "margin:0.4rem 0 0 1rem; padding:0;",
        lapply(packages, function(p) tags$li(p, style = "font-size:0.88rem; line-height:1.2;"))
      )
    )
  })

# Container with grid styles inserted (scoped CSS)
out <- tagList(
  tags$style(HTML("
    .card-grid { display: grid; gap: 0.8rem; align-items: start; }
    /* 1 column (small) */
    .card-grid { grid-template-columns: repeat(1, 1fr); }
    /* 2 columns on medium screens */
    @media (min-width: 768px) {
      .card-grid { grid-template-columns: repeat(2, 1fr); }
    }
    /* 3 columns on large screens / projectors */
    @media (min-width: 992px) {
      .card-grid { grid-template-columns: repeat(3, 1fr); }
    }
    .pkg-card-title { font-weight: 700; font-size: 0.95rem; margin-bottom: 0.35rem; }
    .pkg-card { font-size: 0.9rem; }
    /* optional: make reveal.js fragments a tiny bit smoother */
    .fragment { transition: transform 240ms ease, opacity 240ms ease; }
  ")),
  tags$div(class = "card-grid", cards_html)
)

# render as HTML for Quarto / revealjs
out
```

```{r, results='asis', echo=FALSE, warning=FALSE, message=FALSE}
packages |>
  dplyr::select(package_name, classification, last_version, title, description, website, github, n_releases, first_release) |>
  purrr::pmap(\(package_name, classification, last_version, title, description, website, github, n_releases, first_release) {
    c(
      paste0("## ", package_name, " (", classification, ")"),
      "",
      ". . .",
      "",
      paste0("**", title, "**"),
      "",
      paste0("Current version: **", last_version, "** First release: *", 
             first_release, "* Total releases: *", n_releases, "*"),
      "",
      paste0("Links: [Website](", website, ") [GitHub Repository](", github, ")"),
      "",
      ". . .",
      "",
      paste0("*", description, "*"), 
      ""
    ) |>
      paste0(collapse = "\n")
  }) |>
  paste0(collapse = "\n\n") |>
  cat()
```

## Dependencies

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(igraph)
library(ggraph)
library(tidygraph)
library(tools)
library(jsonlite)
library(dplyr)
library(purrr)
library(patchwork)
library(ggplot2)
library(cranlogs)
library(bsicons)
library(ggdag)

dependencies <- packages |>
  select(package_name, imports) |>
  tidyr::unnest("imports") |>
  rename(from = package_name, to = imports) |>
  mutate(type = "Imports") |>
  union_all(
    packages |>
      select(package_name, suggests) |>
      tidyr::unnest("suggests") |>
      rename(from = package_name, to = suggests) |>
      mutate(type = "Suggests")
  )

plot_dependencies <- function(dependencies) {
  g <- as_tbl_graph(graph_from_data_frame(dependencies, directed = TRUE, vertices = pkgs))
  g <- g %>%
    mutate(
      dependencies = centrality_degree(mode = "in")
    )
  ggraph(g, layout = "centrality", centrality = dependencies) +
    geom_edge_arc(aes(linetype = type, color = type),
                  strength = 0.1, end_cap = circle(4, 'mm'),
                  arrow = arrow(length = unit(3, 'mm'), type = "closed")) +
    geom_node_point(aes(fill = dependencies),
                    shape = 21, colour = "black",
                    size = 5) +
    geom_node_text(aes(label = name),
                   repel = TRUE,
                   bg.color = "white",
                   bg.r = 0.15,
                   size = 4) +
    scale_edge_linetype_manual(name = "Dependency type",
                               values = c("Imports" = "solid",
                                          "Suggests" = "dashed")) +
    scale_edge_color_manual(name = "Dependency type",
                            values = c("Imports" = "black",
                                       "Suggests" = "grey45")) +
    scale_fill_viridis(name = "Reverse dependencies",
                       option="mako",
                       direction = -1) +
    theme_void() +
    theme(legend.position = "bottom",
          legend.box = "veritcal") 
}
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
dependencies |>
  filter(type == "Imports") |>
  plot_dependencies()
```

## Dependencies

```{r, echo=FALSE, message=FALSE, warning=FALSE}
dependencies |>
  plot_dependencies()
```

## Statistics

```{r, echo=FALSE, message=FALSE, warning=FALSE}
downloads <- cranlogs::cran_downloads(
  packages = packages$package_name, 
  from = as.Date("2022-01-01"), 
  to = Sys.Date()
) 

total_per_package <- downloads |> 
  group_by(package) |> 
  summarise(total = sum(count)) |> 
  ungroup()

total <- total_per_package |> 
  summarise(total = sum(total))

bslib::value_box(
  title = "Number of R packages",
  value = nrow(packages),
  showcase = bsicons::bs_icon("bar-chart")
)

bslib::value_box(
  title = "Number of R packages on CRAN",
  value = total_per_package |> nrow(),
  showcase = bsicons::bs_icon("bar-chart")
)

bslib::value_box(
  title = "Number Releases",
  value = versions |> nrow(),
  showcase = bsicons::bs_icon("bar-chart")
)

bslib::value_box(
  title = "Total CRAN downloads",
  value = total |> pull(),
  showcase = bsicons::bs_icon("bar-chart")
)
```

## Outline

-   classes
-   methods
-   Support more back-ends
-   Build a package on top

## Clases

### cdm_reference

### cdm_table

### cdm_source

### omop_table

### achilles_table

### cohort_table

### codelist, codelist_with_details, concept_set_expression

### summarised_result

## Methods

## Support more back-ends

- Help wanted
- Write documentation on how to add more back ends

## Build a package on top

- Tidy R packages
